Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-23 latex/main.tex
Decent pass at para 3

% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
% we lack methods to build phylogenies that can integrate recombination
% in the face of ongoing sampling of infected cases.

% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
% UShER has generated phylogenies updated daily \cite{McBroome2021} and
% has proven to be a powerful resource that successfully enabled the world
% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
% These phylogenies are used to identify and monitor
% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
% which might have acquired features to become Variants of Concern (VOC),
% for example, increased transmissibility, disease severity, and immune escape.
% Beyond supporting real-time genomic epidemiology,
% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
% UShER uses a tree to represent evolutionary change via mutation,
% assuming no recombination \cite{Turakhia2021}.

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024}.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.

Previous ARG methods are retrospective \cite{Rasmussen2014,Kelleher2019,Speidel2019,Ignatieva2022,Zhang2023},
reconstructing the past from a sample of genome sequences drawn at one moment in time,
but scientists are increasingly tracking evolutionary change as it happens,
accumulating data forward in time.
For example, sequences from daily samples of SARS-CoV-2 were made globally available on a daily basis,
leading to over 17 million sequences in the GISAID database \cite{Shu2017}.

Here, we present
(1) sc2ts, a new method to infer ARGs in real time by building genealogies forward in time, and
(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences.
We used the ``Viridian'' dataset,
which is purged of various systematic errors and artifacts \cite{Hunt2024} and
show that this ARG is highly congruent to an UShER phylogeny built from the same dataset.
Compared to previously identified recombinants,
we show that this ARG contains the vast majority of epidemiologically relevant signals of recombination,
while adding significant nuance and texture.
This ARG – which captures the major signals of evolution
through phylogenetic relationships, point mutations, deletions, and recombination –
is a resource of unprecedented detail and precision, and
should provide a strong basis for future studies of SARS-CoV-2.

>>>>

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
of interest in their application in population and statical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
A subtle but important point distinguishing recent work on ARGs
from the classical literature is that the ARG 
``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
tightly coupled to the generative
process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
Another important point is that strong software support for ARGs now
exists, and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing 
software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

[TODO rewrite this paragraph]
Here, we present
(1) sc2ts, a new method to infer ARGs in real time by building genealogies forward in time, and
(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences.
We used the ``Viridian'' dataset,
which is purged of various systematic errors and artifacts \cite{Hunt2024} and
show that this ARG is highly congruent to an UShER phylogeny built from the same dataset.
Compared to previously identified recombinants,
we show that this ARG contains the vast majority of epidemiologically relevant signals of recombination,
while adding significant nuance and texture.
This ARG – which captures the major signals of evolution
through phylogenetic relationships, point mutations, deletions, and recombination –
is a resource of unprecedented detail and precision, and
should provide a strong basis for future studies of SARS-CoV-2.
